# Causal Mediation Analysis {#mediation}

## Mediation models

In this workshopp we will use directed acyclic graphs to conceptualize
the effects of interest. We will focus on the two types of graph:

### No intermediate confounders

```{r, engine="tikz", fig.cap = "Directed acylcic graph under *no intermediate confounders* of the mediator-outcome relation affected by treatment", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", echo = FALSE}
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[2]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
  \Vertex{-4, 0}{W}
  \Vertex{0, 0}{M}
  \Vertex{-2, 0}{A}
  \Vertex{2, 0}{Y}
  \Arrow{W}{A}{black}
  \Arrow{A}{M}{black}
  \Arrow{M}{Y}{black}
  \ArrowL{W}{Y}{black}
  \ArrowL{A}{Y}{black}
  \ArrowL{W}{M}{black}
\end{tikzpicture}
```

### Intermediate confounders

```{r, engine="tikz", fig.cap = "Directed acylcic graph under intermediate confounders of the mediator-outcome relation affected by treatment", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", echo = FALSE}
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[2]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
  \Vertex{0, -1}{Z}
  \Vertex{-4, 0}{W}
  \Vertex{0, 0}{M}
  \Vertex{-2, 0}{A}
  \Vertex{2, 0}{Y}
  \ArrowR{W}{Z}{black}
  \Arrow{Z}{M}{black}
  \Arrow{W}{A}{black}
  \Arrow{A}{M}{black}
  \Arrow{M}{Y}{black}
  \Arrow{A}{Z}{black}
  \Arrow{Z}{Y}{black}
  \ArrowL{W}{Y}{black}
  \ArrowL{A}{Y}{black}
  \ArrowL{W}{M}{black}
\end{tikzpicture}
```

## Counterfactuals

In what follows, we will define all the effects of interest using
_counterfactuals_. Counterfactuals are hypothetical random variables
that would have been observed in a world where we would be able to
perform interventions on the random variables of interest. For
example, $Y_a$ is a counterfactual variable in a hypothetical world
where $\P(A=a)=1$ with probability one. $Y_{a,m}$ is the
counterfactual outcome in a world where $\P(A=a,M=m)=1$, and $M_a$ is
the counterfactual variable representing the mediator in a world where
$\P(A=a)=1$.

In this workshop we use counterfactual variables as _primitives_, but
we note that in other causal inference frameworks, such as structural
equation models, counterfactuals are quantities _derived_ from the
model.
