# How to choose an estimand: Real world example {#estimandirl}

## Comparative effectivness of two medications for opioid use disorder (OUD)

```{r pressure, echo=FALSE, fig.cap="", out.width = '50%'}
knitr::include_graphics(here::here("img/ctndag.pdf"))
```

_Motivation_: Opposite overall treatment effects for homeless versus
nonhomeless participants.

```{r , echo=FALSE, fig.cap="", out.width = '50%'}
knitr::include_graphics(here::here("img/tmleesttotal.pdf"))
```

### Getting specific about the question

To what extent does the indirect effect through mediators of adherence, pain,
and depressive symptoms explain the differences in treatment effects on OUD
relapse for homeless and nonhomeless individuals?

What estimand do we want?

- Can we set $M=m$ (i.e., same value) for everyone?
- Are we interested in estimating indirect effects?

$\rightarrow$ So, _not_ controlled direct effect.

- Do we have an intermediate confounder?
 + Yes, and it's important.

$\rightarrow$ So, _not_ natural (in)direct effects.

- So, we're left with the interventional direct and indirect effects.
- Do we want to estimate the path through treatment initiation ($Z$)?
 + Yes, so, _not_ the conditional versions of these effects.
 + Estimands:
   - Direct effect: $\E(Y_{1,g_0} - Y_{0,g_0})$
   - Indirect effect: $\E(Y_{1,g_1} - Y_{1,g_0})$
- Need to incorporate multiple and continuous mediators

