# `R` packages for estimation of the causal (in)direct effects

We'll now turn to working through a few examples of estimating the natural,
interventional, and stochastic direct and indirect effects. As our running
example, we'll a simple data set from an observational study of the relationship
between BMI and kids' behavior, freely distributed with the [`mma` `R` package
on CRAN](https://CRAN.R-project.org/package=mma). First, let's load the packages
we'll be using and set a seed; then, load this data set and take a quick look

```{r setup}
library(tidyverse)
library(sl3)
library(medoutcon)
library(medshift)
library(mma)
set.seed(429153)
```

```{r load_data}
# load and examine data
data(weight_behavior)
dim(weight_behavior)

# drop missing values
weight_behavior <- weight_behavior %>%
  drop_na() %>%
  as_tibble()
weight_behavior
```

The documentation for the data set describes it as a "database obtained from the
Louisiana State University Health Sciences Center, New Orleans, by Dr. Richard
Scribner. He explored the relationship between BMI and kids' behavior through a
survey at children, teachers and parents in Grenada in 2014. This data set
includes 691 observations and 15 variables." Note that the data set contained
several observations with missing values, which we removed above to simplify the
demonstration of our analytic methods. In practice, we recommend instead using
appropriate corrections (e.g., imputation, inverse weighting) to fully take
advantage of the observed data.

Following the motivation of the original study, we focus on the causal effects
of participating in a sports team (`sports`) on the BMI of children (`bmi`),
taking into consideration several mediators (`snack`, `exercises`, `overweigh`);
all other measured covariates are taken to be potential baseline confounders.

## `medoutcon`: Natural and interventional (in)direct effects

The data on a single observational unit can be represented $O = (W, A, M, Y)$,
with the data pooled across all participants denoted $O_1, \ldots, O_n$, for a
of $n$ i.i.d. observations of $O$. Recall the DAG [from an earlier
chapter](#estimands), which represents the data-generating process:

```{r, engine="tikz", fig.cap = "Directed acyclic graph under *no intermediate confounders* of the mediator-outcome relation affected by treatment", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", echo = FALSE}
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[2]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
  \Vertex{-4, 0}{W}
  \Vertex{0, 0}{M}
  \Vertex{-2, 0}{A}
  \Vertex{2, 0}{Y}
  \Arrow{W}{A}{black}
  \Arrow{A}{M}{black}
  \Arrow{M}{Y}{black}
  \ArrowL{W}{Y}{black}
  \ArrowL{A}{Y}{black}
  \ArrowL{W}{M}{black}
\end{tikzpicture}
```

### Natural (in)direct effects

To start, we will consider estimation of the _natural_ direct and indirect effects,
which, we recall, are defined as follows
\begin{equation*}
  \E[Y_{1,M_1} - Y_{0,M_0}] = \underbrace{\E[Y_{\color{red}{1},\color{blue}{M_1}} -
    Y_{\color{red}{1},\color{blue}{M_0}}]}_{\text{natural indirect effect}} +
    \underbrace{\E[Y_{\color{blue}{1},\color{red}{M_0}} -
    Y_{\color{blue}{0},\color{red}{M_0}}]}_{\text{natural direct effect}}.
\end{equation*}

* Our [`medoutcon` `R` package](https://github.com/nhejazi/medoutcon)
  [@hejazi2021medoutcon], which accompanies @diaz2020nonparametric, implements
  one-step and TML estimators of both the natural and interventional (in)direct
  effects.
* Both types of estimators are capable of accommodating flexible modeling
  strategies (e.g., ensemble machine learning) for the initial estimation of
  nuisance parameters.
* The `medoutcon` `R` package uses cross-validation in initial estimation: this
  results in cross-validated (or "cross-fitted") one-step and TML estimators
  [@klaassen1987consistent; @zheng2011cross; @chernozhukov2018double], which
  exhibit greater robustness than their non-sample-splitting analogs.
* To this end, `medoutcon` integrates with the `sl3` `R` package, which is
  extensively documented in this [book
  chapter](https://tlverse.org/tlverse-handbook/sl3) [@vdl2022targeted].

### Interlude: `sl3` for nuisance parameter estimation

* To fully take advantage of the one-step and TML estimators, we'd like to rely
  on flexible, data adaptive strategies for nuisance parameter estimation.
* Doing so minimizes opportunities for model misspecification to compromise our
  analytic conclusions.
* Choosing among the diversity of available machine learning algorithms can be
  challenging, so we recommend using the Super Learner algorithm for ensemble
  machine learning [@vdl2007super], which is implemented in the [`sl3` R
  package](https://github.com/tlverse/sl3) [@coyle2021sl3].
* Below, we demonstrate the construction of an ensemble learner based on a
  limited library of algorithms, including n intercept model, a main terms GLM,
  Lasso ($\ell_1$-penalized) regression, and random forest (`ranger`).
```{r setup_sl}
# instantiate learners
mean_lrnr <- Lrnr_mean$new()
fglm_lrnr <- Lrnr_glm_fast$new()
lasso_lrnr <- Lrnr_glmnet$new(alpha = 1, nfolds = 3)
rf_lrnr <- Lrnr_ranger$new(num.trees = 200)

# create learner library and instantiate super learner ensemble
lrnr_lib <- Stack$new(mean_lrnr, fglm_lrnr, lasso_lrnr, rf_lrnr)
sl_lrnr <- Lrnr_sl$new(learners = lrnr_lib, metalearner = Lrnr_nnls$new())
```

* Of course, there are many alternatives for learning algorithms to be included
  in such a modeling library. Feel free to explore!

### Efficient estimation of the natural (in)direct effects

* Estimation of the natural direct and indirect effects requires estimation of a
  few nuisance parameters. Recall that these are
  - $g(a\mid w)$, which denotes $\P(A=a \mid W=w)$
  - $h(a\mid m, w)$, which denotes $\P(A=a \mid M=m, W=w)$
  - $b(a, m, w)$, which denotes $\E(Y \mid A=a, M=m, W=w)$
* While we recommend the use of Super Learning, we opt to instead estimate all
  nuisance parameters with Lasso regression below (to save computational time).
* Now, we're ready to use the `medoutcon` function to estimate the _natural
  direct effect_:
```{r natural_de_os}
# compute one-step estimate of the natural direct effect
nde_onestep <- medoutcon(
  W = weight_behavior[, c("age", "sex", "race", "tvhours")],
  A = (as.numeric(weight_behavior$sports) - 1),
  Z = NULL,
  M = weight_behavior[, c("snack", "exercises", "overweigh")],
  Y = weight_behavior$bmi,
  g_learners = lasso_lrnr,
  h_learners = lasso_lrnr,
  b_learners = lasso_lrnr,
  effect = "direct",
  estimator = "onestep",
  estimator_args = list(cv_folds = 5)
)
summary(nde_onestep)
```

* We can similarly call the `medoutcon` function to estimate the _natural
  indirect effect_:
```{r natural_ie_os}
# compute one-step estimate of the natural indirect effect
nie_onestep <- medoutcon(
  W = weight_behavior[, c("age", "sex", "race", "tvhours")],
  A = (as.numeric(weight_behavior$sports) - 1),
  Z = NULL,
  M = weight_behavior[, c("snack", "exercises", "overweigh")],
  Y = weight_behavior$bmi,
  g_learners = lasso_lrnr,
  h_learners = lasso_lrnr,
  b_learners = lasso_lrnr,
  effect = "indirect",
  estimator = "onestep",
  estimator_args = list(cv_folds = 5)
)
summary(nie_onestep)
```

* From the above, we can conclude that the effect of participation on a sports
  team on BMI is primarily mediated by the variables `snack`, `exercises`, and
  `overweigh`, as the natural indirect effect is several times larger than the
  natural direct effect.
* Note that we could have instead used the TML estimators, which have improved
  finite-sample performance, instead of the one-step estimators. Doing this is
  as simple as setting the `estimator = "tmle"` in the relevant argument.

### Interventional (in)direct effects

Since our knowledge of the system under study is incomplete, we might worry that
one (or more) of the measured variables are not mediators, but, in fact,
intermediate confounders affected by treatment. While the natural (in)direct
effects are not identified in this setting, their interventional (in)direct
counterparts are, as we saw in an earlier section. Recall that both types of
effects are defined by static interventions on the treatment. The interventional
effects are distinguished by their use of a stochastic intervention on the
mediator to aid in their identification.

  ```{r, engine="tikz", fig.cap = "Directed acyclic graph under intermediate confounders of the mediator-outcome relation affected by treatment", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", echo = FALSE}
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[2]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
  \Vertex{0, -1}{Z}
  \Vertex{-4, 0}{W}
  \Vertex{0, 0}{M}
  \Vertex{-2, 0}{A}
  \Vertex{2, 0}{Y}
  \ArrowR{W}{Z}{black}
  \Arrow{Z}{M}{black}
  \Arrow{W}{A}{black}
  \Arrow{A}{M}{black}
  \Arrow{M}{Y}{black}
  \Arrow{A}{Z}{black}
  \Arrow{Z}{Y}{black}
  \ArrowL{W}{Y}{black}
  \ArrowL{A}{Y}{black}
  \ArrowL{W}{M}{black}
\end{tikzpicture}
```

Recall that the interventional (in)direct effects are defined via the decomposition:
\begin{equation*}
\E[Y_{1,G_1} - Y_{0,G_0}] = \underbrace{\E[Y_{\color{red}{1},\color{blue}{G_1}} -
    Y_{\color{red}{1},\color{blue}{G_0}}]}_{\text{interventional indirect effect}} +
    \underbrace{\E[Y_{\color{blue}{1},\color{red}{G_0}} -
    Y_{\color{blue}{0},\color{red}{G_0}}]}_{\text{interventional direct effect}}
\end{equation*}

* In our data example, we'll consider the eating of snacks as a potential
  intermediate confounder, since one might reasonably hypothesize that
  participation on a sports team might subsequently affect snacking, which then
  could affect mediators like the amount of exercises and overweight status.
* The interventional direct and indirect effects may also be easily estimated
  with the [`medoutcon` `R` package](https://github.com/nhejazi/medoutcon).
* Just as for the natural (in)direct effects, `medoutcon` implements
  cross-validated one-step and TML estimators of the interventional effects.

### Efficient estimation of the interventional (in)direct effects

* Estimation of these effects is more complex, so a few additional nuisance
  parameters arise when expressing the (more general) EIF for these effects:
  * $q(z \mid a, w)$, the conditional density of the intermediate confounders,
    conditional only on treatment and baseline covariates;
  * $r(z \mid a, m, w)$, the conditional density of the intermediate
    confounders, conditional on mediators, treatment, and baseline covariates.
* Note that the implementation in `medoutcon` is currently limited to settings
  with only binary intermediate confounders, i.e., $Z \in \{0, 1\}$.
* Now, we're ready to use the `medoutcon` function to estimate the
  _interventional direct effect_:
```{r interv_de_os}
# compute one-step estimate of the interventional direct effect
interv_de_onestep <- medoutcon(
  W = weight_behavior[, c("age", "sex", "race", "tvhours")],
  A = (as.numeric(weight_behavior$sports) - 1),
  Z = (as.numeric(weight_behavior$snack) - 1),
  M = weight_behavior[, c("exercises", "overweigh")],
  Y = weight_behavior$bmi,
  g_learners = lasso_lrnr,
  h_learners = lasso_lrnr,
  b_learners = lasso_lrnr,
  effect = "direct",
  estimator = "onestep",
  estimator_args = list(cv_folds = 5)
)
summary(interv_de_onestep)
```

* We can similarly call the `medoutcon` function to estimate the
  _interventional indirect effect_:
```{r interv_ie_os}
# compute one-step estimate of the interventional indirect effect
interv_ie_onestep <- medoutcon(
  W = weight_behavior[, c("age", "sex", "race", "tvhours")],
  A = (as.numeric(weight_behavior$sports) - 1),
  Z = (as.numeric(weight_behavior$snack) - 1),
  M = weight_behavior[, c("exercises", "overweigh")],
  Y = weight_behavior$bmi,
  g_learners = lasso_lrnr,
  h_learners = lasso_lrnr,
  b_learners = lasso_lrnr,
  effect = "indirect",
  estimator = "onestep",
  estimator_args = list(cv_folds = 5)
)
summary(interv_ie_onestep)
```

* From the above, we can conclude that the effect of participation on a sports
  team on BMI is largely through the interventional indirect effect (i.e.,
  through the pathways involving the mediating variables) rather than via its
  direct effect.
* Just as before, we could have instead used the TML estimators, instead of the
  one-step estimators. Doing this is as simple as setting the
  `estimator = "tmle"` in the relevant argument.

## `medshift`: Stochastic (in)direct effects

While the analyses using the natural and interventional effects have been
illuminating, we may also go beyond the restrictive static interventions
required to define these (in)direct effects. In fact, it may be more realistic
to consider interventions that do not directly force children to join athletic
teams, but instead motivate them to make their participation on such teams more
likely. Importantly, such interventions are often far more realistic and
actionable in real-world studies.

### Formulating the stochastic (in)direct effects

* These more flexible intervention regimes are incompatible with (in)direct
  effect definitions based on decomposing the average treatment effect.
* Instead, consider the decomposition of the population intervention effect
  (PIE) of a _stochastic intervention_ into direct and indirect effects
  [@diaz2020causal]:
\begin{align*}
  \E[Y&_{A_\delta,M_{A_\delta}} - Y_{A,M_A}] = \\
  &\underbrace{\E[Y_{\color{red}{A_\delta},\color{blue}{M_{A_\delta}}} -
    Y_{\color{red}{A_\delta},\color{blue}{M}}]}_{\text{stochastic natural indirect effect}} +
    \underbrace{\E[Y_{\color{blue}{A_\delta},\color{red}{M}} -
    Y_{\color{blue}{A},\color{red}{M}}]}_{\text{stochastic natural direct effect}}
\end{align*}
* Recall from our discussion of the [incremental propensity score
  interventions](#ipsi) [@kennedy2018nonparametric] that such stochastic
  interventions can compare the pre- and post-intervention odds of exposure:
\begin{equation*}
  \delta = \frac{\text{odds}(A_\delta = 1\mid W=w)}
  {\text{odds}(A = 1\mid W=w)}.
\end{equation*}
* In our analysis, we will modulate the _odds of participating in a sports
  team_ by a fixed amount for each individual, setting, for example,
  $\delta = 2$:
```{r delta_ipsi}
delta_shift_ipsi <- 2
```
* Such an intervention may be interpreted as the effect of a school program that
  motivates children to participate in sports teams.

### Efficient estimation of the stochastic (in)direct effects

* The decomposition of the PIE into the direct and indirect effects leads to a
  common term $\E[Y_{\color{red}{A_\delta},\color{blue}{M}}]$ involved in both
  the direct and indirect effect definitions. This term may be estimated via the
  [`medshift` `R` package](https://github.com/nhejazi/medshift)
  [@hejazi2020medshift].
* For the direct effect, the remaining term is the
  $\E[Y_{\color{blue}{A},\color{red}{M}}]$, which may be estimated by a simple
  mean in the observed data (i.e., no intervention).
* For the indirect effect, the remaining term is the joint effect of stochastic
  interventions on both $A$ and $M$:
  $\E[Y_{\color{red}{A_\delta},\color{blue}{M_{A_\delta}}}$.
  * For the case of an IPSI on binary $A$, this may be estimated by the tools
    in the [`npcausal` `R` package](https://github.com/ehkennedy/npcausal).
  * For the case of an MTP on continuous $A$, this may be estimated by the tools
    in the [`txshift` `R` package](https://github.com/nhejazi/txshift)
    [@hejazi2020txshift-rpkg; @hejazi2020txshift-joss].
* Like the implementation in `medoutcon`, the `medshift` package makes use of
  cross-validation in constructing initial estimates of nuisance parameters,
  resulting in more robust, cross-validated efficient estimators
  [@klaassen1987consistent; @zheng2011cross; @chernozhukov2018double].
* Now, we're ready to use the `medshift` function to estimate the decomposition
  term common to both the _stochastic direct and indirect effects_:
```{r interv_de_os}
# compute one-step estimate of the decomposition term of the (in)direct effects
stoch_decomp_onestep <- medshift(
  W = weight_behavior[, c("age", "sex", "race", "tvhours")],
  A = (as.numeric(weight_behavior$sports) - 1),
  Z = weight_behavior[, c("snack", "exercises", "overweigh")],
  Y = weight_behavior$bmi,
  delta = delta_shift_ipsi,
  g_learners = lasso_lrnr,
  e_learners = lasso_lrnr,
  m_learners = lasso_lrnr,
  estimator = "onestep",
  estimator_args = list(cv_folds = 5)
)
summary(stoch_decomp_onestep)
```
<!-- Nima: note the notational differences from the above -->

* To estimate the stochastic direct effect, an extra step is necessary -- we
  must apply the delta method:
```{r linear_contrast_delta}
# convenience function to compute inference via delta method: EY1 - EY0
linear_contrast <- function(params, eifs, ci_level = 0.95) {
  # bounds for confidence interval
  ci_norm_bounds <- c(-1, 1) * abs(stats::qnorm(p = (1 - ci_level) / 2))
  param_est <- params[[1]] - params[[2]]
  eif <- eifs[[1]] - eifs[[2]]
  se_eif <- sqrt(var(eif) / length(eif))
  param_ci <- param_est + ci_norm_bounds * se_eif
  # parameter and inference
  out <- c(param_ci[1], param_est, param_ci[2])
  names(out) <- c("lwr_ci", "param_est", "upr_ci")
  return(out)
}
```
* Straightforward application of this procedure yields,
```{r comp_de_binary}
# parameter estimates and EIFs for components of direct effect
EY <- mean(weight_behavior$bmi)
eif_EY <- weight_behavior$bmi - EY
params_de <- list(stoch_decomp_onestep$theta, EY)
eifs_de <- list(stoch_decomp_onestep$theta, eif_EY)

# direct effect = EY - estimated quantity
de_est <- linear_contrast(params_de, eifs_de)
de_est
```
* From the above, we can conclude that the effect of increasing the odds of
  participation on a sports team on BMI leads only to a relatively small direct
  effect.
