# Super (Machine) Learning {#sl3}

_Rachael Phillips_

Based on the [`sl3` `R` package](https://github.com/tlverse/sl3) by _Jeremy
Coyle, Nima Hejazi, Ivana Malenica, Rachael Phillips, and Oleg Sofrygin_.

Updated: `r Sys.Date()`

## Learning Objectives {-}

By the end of this chapter you will be able to:

1. Select a loss function that is appropriate for the functional parameter to be
   estimated.
2. Assemble an ensemble of learners based on the properties that identify what
   features they support.
3. Customize learner hyperparameters to incorporate a diversity of different
   settings.
4. Select a subset of available covariates and pass only those variables to the
   modeling algorithm.
5. Fit an ensemble with nested cross-validation to obtain an estimate of the
   performance of the ensemble itself.
6. Obtain `sl3` variable importance metrics.
7. Interpret the discrete and continuous Super Learner fits.
8. Rationalize the need to remove bias from the Super Learner to make an optimal
   bias–variance tradeoff for the parameter of interest.

## Motivation {-}

- A common task in statistical data analysis is estimator selection (e.g., for
  prediction).
- There is no universally optimal machine learning algorithm for density
  estimation or prediction.
- For some data, one needs learners that can model a complex function.
- For others, possibly as a result of noise or insufficient sample size, a
  simple, parametric model might fit best.
- The Super Learner, an ensemble learner, solves this issue, by allowing a
  combination of learners from the simplest (intercept-only) to most complex
  (neural nets, random forests, SVM, etc).
- It works by using cross-validation in a manner which guarantees that the
  resulting fit will be as good as possible, given the learners provided.

## Introduction {-}

In [Chapter 1](#intro), we introduced the [_Roadmap for Targeted
Learning_](#roadmap) as a general template to translate real-world data
applications into formal statistical estimation problems. The first steps of
this roadmap define the *statistical estimation problem*, which establish

1. Data as a realization of a random variable, or equivalently, an outcome of a
   particular experiment.
2. A statistical model, representing the true knowledge about the
   data-generating experiment.
3. A translation of the scientific question, which is often causal, into a
   target parameter.

Note that if the target parameter is causal, step 3 also requires
establishing identifiability of the target quantity from the observed data
distribution, under possible non-testable assumptions that may not necessarily
be reasonable. Still, the target quantity does have a valid statistical
interpretation. See [causal target parameters](#causal) for more detail on
causal models and identifiability.

Now that we have defined the statistical estimation problem, we are ready to
construct the TMLE; an asymptotically linear and efficient substitution
estimator of this target quantity. The first step in this estimation procedure
is an initial estimate of the data-generating distribution, or the relevant part
of this distribution that is needed to evaluate the target parameter. For this
initial estimation, we use the Super Learner [@vdl2007super].

The Super Learner provides an important step in creating a robust estimator. It
is a loss-function-based tool that uses cross-validation to obtain the best
prediction of our target parameter, based on a weighted average of a library of
machine learning algorithms.

The library of machine learning algorithms consists of functions ("learners" in
the `sl3` nomenclature) that we think might be consistent with the true
data-generating distribution (i.e. algorithms selected based on contextual
knowledge of the experiment that generated the data). Also, the library should
contain a large set of "default" algorithms that may range from a simple linear
regression model to multi-step algorithms involving screening covariates,
penalizations, optimizing tuning parameters, etc.

The ensembling of the collection of algorithms with weights ("metalearning" in
the `sl3` nomenclature) has been shown to be adaptive and robust, even in small
samples [@polley2010super]. The Super Learner is proven to be asymptotically as
accurate as the best possible prediction algorithm in the library
[@vdl2003unified; @vaart2006oracle].

### Background {-}

A *loss function* $L$ is defined as a function of the observed data and a
candidate parameter value $\psi$, which has unknown true value $\psi_0$,
$L(\psi)(O)$. We can estimate the loss by substituting the empirical
distribution $P_n$ for the true (but unknown) distribution of the observed data
$P_0$. A valid loss function will have expectation (risk) that is minimized at
the true value of the parameter $\psi_0$. For example, the conditional mean
minimizes the risk of the squared error loss. Thus, it is a valid loss function
when estimating the conditional mean.

The *discrete Super Learner*, or *cross-validation selector*, is the algorithm
in the library that minimizes the cross-validated empirical risk.

The *cross-validated empirical risk* of an algorithm is defined as the empirical
mean over a validation sample of the loss of the algorithm fitted on the
training sample, averaged across the splits of the data.

The *continuous/ensemble Super Learner*, often referred to as *Super Learner*
is a weighted average of the library of algorithms, where the weights are chosen
to minimize the cross-validated empirical risk of the library. Restricting the
weights to be positive and sum to one (i.e., a convex combination) has been
shown to improve upon the discrete Super Learner [@polley2010super;
@vdl2007super]. This notion of weighted combinations was introduced in
@wolpert1992stacked for neural networks and adapted for regressions in
@breiman1996stacked.

Cross-validation is proven to be optimal for selection among estimators. This
result was established through the oracle inequality for the cross-validation
selector among a collection of candidate estimators [@vdl2003unified;
@vaart2006oracle]. The only condition is that loss function is uniformly
bounded, which is guaranteed in `sl3`.

<!--
The *oracle results* prove that, if the number of algorithms in the library are
polynomial in sample size, then the cross-validation selector (i.e., discrete
Super Learner) (1) is equivalent with the oracle selector asymptotically (based
on sample of size of training samples), or (2) achieves the parametric rate
(log $n/n$) for convergence with respect to the loss-based dissimilarity (risk)
between a candidate estimate $\psi$ and the true parameter value $\psi_0$.
-->

#### General Overview of the Algorithm {-}

**What is cross-validation and how does it work?**

There are many different cross-validation schemes, designed to accommodate
different study designs and data structures. The figure below shows an example
of 10-fold cross-validation.   

```{r cv_fig, results="asis", echo=FALSE}
knitr::include_graphics("img/misc/vs.pdf")
```

**General step-by-step overview of the Super Learner algorithm:**

- Break up the sample evenly into V-folds (say V=10).
- For each of these 10 folds, remove that portion of the sample (kept out as
  validation sample) and the remaining will be used to fit learners (training
  sample).
- Fit each learner on the training sample (note, some learners will have their
  own internal cross-validation procedure or other methods to select tuning
  parameters).
- For each observation in the corresponding validation sample, predict the outcome
  using each of the learners, so if there are $p$ learners, then there would be
  $p$ predictions.
- Take out another validation sample and repeat until each of the V-sets of data
  are removed.
- Compare the cross-validated fit of the learners across all observations based
  on specified loss function (e.g., squared error, negative log-likelihood, ...)
  by calculating the corresponding average loss (risk).
- Either:

  + choose the learner with smallest risk and apply that learner to entire data
    set (resulting SL fit),
  + do a weighted average of the learners to minimize the cross-validated risk
    (construct an ensemble of learners), by

    + re-fitting the learners on the original data set, and
    + use the weights above to get the SL fit.

This entire procedure can be itself cross-validated to get a consistent
estimate of the future performance of the Super Learner, and we implement this
procedure later in this chapter.

```{r cv_fig2, results="asis", echo=FALSE}
knitr::include_graphics("img/misc/SLKaiserNew.pdf")
```

### Super Learner for Prediction {-}

Say we observe a learning data set $X_i=(Y_i,W_i)$, for $i=1, ..., n$, where
$Y_i$ is the outcome of interest, $W_i$ is a $p$-dimensional set of
covariates, and our objective is to estimate the function $\psi_0(W) = E(Y|W)$.
This function can be expressed as the minimizer of the expected loss:
$\psi_0(W) = \text{argmin}_{\psi} E[L(X,\psi(W))]$. Here, the loss function is
represented as $L$ (e.g., squared error loss, $L: (Y-\psi(W))^2)$).

For prediction, one can use the cross-validated risk to empirically determine
the relative performance of the Super Learner. When we have tested different
algorithms on actual data and looked at the performance (e.g., MSE of
prediction), never does one algorithm always win. Below, we show the results of
such a study, comparing the fits of several different learners, including the SL
algorithms.

```{r cv_fig3, results="asis", echo = FALSE}
knitr::include_graphics("img/misc/ericSL.pdf")
```

For more detail on Super Learner we refer the reader to @vdl2007super and
@polley2010super. The optimality results for the cross-validation selector
among a family of algorithms were established in @vdl2003unified and extended
in @vaart2006oracle.

## `sl3` "Microwave Dinner" Implementation {-}

We begin by illustrating the core functionality of the super learner algorithm
as implemented in `sl3`. For those who are interested in the internals of `sl3`,
see this [`sl3` introductory
tutorial](https://tlverse.org/sl3/articles/intro_sl3.html).

The `sl3` implementation consists of the following steps:

0. Load the necessary libraries and data
1. Define the machine learning task
2. Make a super learner by creating library of base learners and a metalearner
3. Train the super learner on the machine learning task
4. Obtain predicted values

### WASH Benefits Study Example {-}

Using the WASH data, we are interested in predicting weight-for-height z-score
`whz` using the available covariate data. More information on this dataset, and
all other data that we will work with in this handbook, is contained in [Chapter
3]{#data}. Let's begin!

### 0. Load the necessary libraries and data {-}

First, we will load the relevant `R` packages, set a seed, and load the data.

<!--
If you would like to use newer `sl3` functionality that is available in the
devel branch of the `sl3` GitHub repository, you need to install that version
of the package (i.e., `usethis::install_github(tlverse/sl3@devel)`), re-start
your `R` session, and then re-load the `sl3` package.
-->

```{r setup}
library(data.table)
library(dplyr)
library(readr)
library(ggplot2)
library(SuperLearner)
library(origami)
library(sl3)
library(knitr)
library(kableExtra)

# load data set and take a peek
washb_data <- fread(
  paste0(
    "https://raw.githubusercontent.com/tlverse/tlverse-data/master/",
    "wash-benefits/washb_data.csv"
  ),
  stringsAsFactors = TRUE
)
head(washb_data) %>%
  kable() %>%
  kableExtra:::kable_styling(fixed_thead = T) %>%
  scroll_box(width = "100%", height = "300px")
```

### 1. Define the machine learning task {-}

To define the machine learning **"task"** (predict weight-for-height Z-score
`whz` using the available covariate data), we need to create an `sl3_Task`
object.

The `sl3_Task` keeps track of the roles the variables play in the machine
learning problem, the data, and any metadata (e.g., observational-level weights,
IDs, offset).

Also, if we had missing outcomes, we would need to set `drop_missing_outcome =
TRUE` when we create the task. In the next analysis, with the [IST stroke trial
data](#ist), we do have a missing outcome. In the following chapter, we estimate
the missingness mechanism and account for it in the TMLE.

```{r task}
# specify the outcome and covariates
outcome <- "whz"
covars <- colnames(washb_data)[-which(names(washb_data) == outcome)]

# create the sl3 task
washb_task <- make_sl3_Task(
  data = washb_data,
  covariates = covars,
  outcome = outcome
)
```
*This warning is important.* The task just imputed missing covariates for us.
Specifically, for each covariate column with missing values, `sl3` uses the
median to impute missing continuous covariates, and the mode to impute binary
and categorical covariates.

Also, for each covariate column with missing values, `sl3` adds an additional
column indicating whether or not the value was imputed, which is particularly
handy when the missingness in the data might be informative.

Also, notice that we did not specify the number of folds, or the loss function
in the task. The default cross-validation scheme is V-fold, with the number of
folds $V=10$.

Let's visualize our `washb_task`.

```{r task-examine}
washb_task
```

### 2. Make a Super Learner {-}

Now that we have defined our machine learning problem with the task, we are
ready to **"make"** the Super Learner. This requires specification of

* A library of base learning algorithms that we think might be consistent with
  the true data-generating distribution.
* A metalearner, to ensemble the base learners.

We might also incorporate

* Feature selection, to pass only a subset of the predictors to the algorithm.
* Hyperparameter specification, to tune base learners.

Learners have properties that indicate what features they support. We may use
`sl3_list_properties()` to get a list of all properties supported by at least
one learner.

```{r list-properties}
sl3_list_properties()
```
Since we have a continuous outcome, we may identify the learners that support
this outcome type with `sl3_list_learners()`.

```{r list-learners}
sl3_list_learners("continuous")
```

Now that we have an idea of some learners, we can construct them using the
`make_learner` function.

```{r baselearners}
# choose base learners
lrnr_glm <- make_learner(Lrnr_glm)
lrnr_mean <- make_learner(Lrnr_mean)
```
We can customize learner hyperparameters to incorporate a diversity of different
settings. Documentation for the learners and their hyperparameters can be found
in the [`sl3` Learners
Reference](https://tlverse.org/sl3/reference/index.html#section-sl-learners).

```{r extra-lrnr-awesome}
lrnr_ranger50 <- make_learner(Lrnr_ranger, num.trees = 50)
lrnr_hal_simple <- make_learner(Lrnr_hal9001, max_degree = 2, n_folds = 2)
lrnr_lasso <- make_learner(Lrnr_glmnet) # alpha default is 1
lrnr_ridge <- make_learner(Lrnr_glmnet, alpha = 0)
lrnr_elasticnet <- make_learner(Lrnr_glmnet, alpha = 0.5)
```

We can also include learners from the `SuperLearner` `R` package.

```{r extra-lrnr-woah}
lrnr_bayesglm <- Lrnr_pkg_SuperLearner$new("SL.bayesglm")
```

Here is a fun trick to create customized learners over a grid of parameters.

```{r extra-lrnr-mindblown-svm, eval = FALSE}
# I like to crock pot my super learners
grid_params <- list(
  cost = c(0.01, 0.1, 1, 10, 100, 1000),
  gamma = c(0.001, 0.01, 0.1, 1),
  kernel = c("polynomial", "radial", "sigmoid"),
  degree = c(1, 2, 3)
)
grid <- expand.grid(grid_params, KEEP.OUT.ATTRS = FALSE)
params_default <- list(nthread = getOption("sl.cores.learners", 1))
svm_learners <- apply(grid, MARGIN = 1, function(params_tune) {
  do.call(Lrnr_svm$new, c(params_default, as.list(params_tune)))
})
```
```{r extra-lrnr-mindblown-xgboost}
grid_params <- list(
  max_depth = c(2, 4, 6, 8),
  eta = c(0.001, 0.01, 0.1, 0.2, 0.3),
  nrounds = c(20, 50)
)
grid <- expand.grid(grid_params, KEEP.OUT.ATTRS = FALSE)
params_default <- list(nthread = getOption("sl.cores.learners", 1))
xgb_learners <- apply(grid, MARGIN = 1, function(params_tune) {
  do.call(Lrnr_xgboost$new, c(params_default, as.list(params_tune)))
})
```

Did you see `Lrnr_caret` when we called `sl3_list_learners(c("binomial"))`?
All we need to specify is the algorithm to use, which is passed as `method` to
`caret::train()`. The default method for parameter selection criterion with
is set to "CV" instead of the `caret::train()` default `boot`. The summary
metric to used to select the optimal model is `RMSE` for continuous outcomes
and `Accuracy` for categorical and binomial outcomes.

```{r carotene, eval = FALSE}
# I have no idea how to tune a neural net (or BART machine..)
lrnr_caret_nnet <- make_learner(Lrnr_caret, algorithm = "nnet")
lrnr_caret_bartMachine <- make_learner(Lrnr_caret,
  algorithm = "bartMachine",
  method = "boot", metric = "Accuracy",
  tuneLength = 10
)
```

In order to assemble the library of learners, we need to **"stack"** them
together.

A `Stack` is a special learner and it has the same interface as all other
learners. What makes a stack special is that it combines multiple learners by
training them simultaneously, so that their predictions can be either combined
or compared.

```{r stack}
stack <- make_learner(
  Stack,
  lrnr_glm, lrnr_mean, lrnr_ridge, lrnr_lasso, xgb_learners[[10]]
)
```

We can optionally select a subset of available covariates and pass only
those variables to the modeling algorithm.

Let's consider screening covariates based on their `randomForest` variable
importance ranking (ordered by mean decrease in accuracy). To select the top
$5$ most important covariates according to this ranking, we can combine
`Lrnr_screener_importance` with `Lrnr_randomForest` (limiting the number of
trees by setting `ntree = 20`).

Hang on! Before you think it -- I will confess: Bob Ross and I both know that 20
trees makes for a lonely forest, and I shouldn't even consider it, but these are
the sacrifices I have to make for this chapter to build in under 50 minutes!

```{r screener}
screen_rf <- Lrnr_screener_importance$new(
  Lrnr_randomForest$new(ntree = 20),
  num_screen = 5
)
# which covariates are selected on the full data?
screen_rf$train(washb_task)
```

To **"pipe"** only the selected covariates to the modeling algorithm, we need to
make a `Pipeline`, which is a just set of learners to be fit sequentially, where
the fit from one learner is used to define the task for the next learner.

```{r screener-pipe}
screen_rf_pipeline <- make_learner(Pipeline, screen_rf, stack)
```
Now our learners will be preceded by a screening step.

We also consider the original `stack`, to compare how the feature selection
methods perform in comparison to the methods without feature selection, and
because

Analogous to what we have seen before, we have to stack the pipeline and
original `stack` together, so we may use them as base learners in our super
learner.

```{r screeners-stack}
fancy_stack <- make_learner(Stack, screen_rf_pipeline, stack)
# we can visualize the stack
dt_stack <- delayed_learner_train(fancy_stack, washb_task)
plot(dt_stack, color = FALSE, height = "400px", width = "90%")
```

We will use the [default
metalearner](https://github.com/tlverse/sl3/blob/master/R/default_metalearner.R),
which uses
[`Lrnr_solnp()`](https://github.com/tlverse/sl3/blob/master/R/Lrnr_solnp.R) to
provide fitting procedures for a pairing of [loss
function](https://github.com/tlverse/sl3/blob/master/R/loss_functions.R) and
[metalearner
function](https://github.com/tlverse/sl3/blob/master/R/metalearners.R). This
default metalearner selects a loss and metalearner pairing based on the outcome
type. Note that any learner can be used as a metalearner.

We have made a library/stack of base learners, so we are ready to make the super
learner. The super learner algorithm fits a metalearner on the validation-set
predictions.

```{r make-sl}
sl <- make_learner(Lrnr_sl,
  learners = fancy_stack
)
```
We can also use `Lrnr_cv` to build a super learner, cross-validate a stack of
learners to compare performance of the learners in the stack, or cross-validate
any single learner (see "Cross-validation" section of this [`sl3`
introductory tutorial](https://tlverse.org/sl3/articles/intro_sl3.html)).

Furthermore, we can [Define New `sl3`
Learners](https://tlverse.org/sl3/articles/custom_lrnrs.html) which can be used
in all the places you could otherwise use any other `sl3` learners, including
`Pipelines`, `Stacks`, and the Super Learner.

In the plot below, we visualize the steps for executing the Super Learner in the
`tlverse/delayed` framework. For those like myself who are not particularly
keen on understanding the intricacies of `delayed`, let's focus on the main
point of this figure: we can see there are 10 realizations of the stack which
represent the 10 cross-validation folds and there is a separate hold-out
(top branch of the figure) that will not be used to fit the Super Learner.

```{r make-sl-plot}
dt_sl <- delayed_learner_train(sl, washb_task)
plot(dt_sl, color = FALSE, height = "400px", width = "90%")
```

### 3. Train the Super Learner on the machine learning task {-}

The Super Learner algorithm fits a metalearner on the validation-set
predictions in a cross-validated manner, thereby avoiding overfitting.

Now we are ready to **"train"** our Super Learner on our `sl3_task` object,
`washb_task`.

```{r sl}
sl_fit <- sl$train(washb_task)
```

### 4. Obtain predicted values {-}

Now that we have fit the super learner, we are ready to calculate the predicted
outcome for each subject.

```{r sl-predictions}
# we did it! now we have super learner predictions
sl_preds <- sl_fit$predict()
head(sl_preds)
```
<!--
Below we visualize the observed versus predicted values.

For fun, we will also
include the cross-validated predictions from most popular learner on the block,
main terms linear regression.


```{r, plot-predvobs-woohoo, eval=FALSE}

df_plot <- data.frame(
  Observed = washb_data$whz,
  Predicted = sl_preds,
  count = c(1:nrow(washb_data))
)

df_plot_melted <- melt(df_plot,
  id.vars = "count",
  measure.vars = c("Observed", "Predicted")
)

ggplot(df_plot_melted, aes(value, count, color = variable)) +
  geom_point()
```
-->

We can also obtain a summary of the results.

```{r, sl-summary}
sl_fit_summary <- sl_fit$print()
```
From the table of the printed Super Learner fit, we note that the Super Learner
had a mean risk of `r sl_fit_summary$mean_risk[length(sl_fit_summary$learner)]`
and that this ensemble weighted the `ranger` and `glmnet` learners highest while
not weighting the `mean` learner highly.

We can also see that the `glmnet` learner had the lowest cross-validated mean
risk, thus making it the cross-validated selector (or the _discrete_ Super
Learner). The mean risk of the Super Learner is calculated using the hold-out
set that we visualized in the `dt_sl` plot.

